---
title: "Working with Geospatial Data in R"
author: "Ian Cook"
date: "June 18, 2015"
output:
  ioslides_presentation:
    logo: Triangle-useR_Logo.png
---

```{r setup, include=FALSE}
options(width=65)
```

## Overview

- Bringing Geospatial Tasks into the Analytics Mainstream
- Geospatial Data Wrangling with R
- Geospatial Analysis with R

# Bringing Geospatial Tasks into the Analytics Mainstream

## Location data

### Organizations capture vast amounts of location data
- Facility locations, customers, suppliers, third-party demographic data, etc.
- ZIP/postal code, address, longitude/latitude

### Location data can have great value if put to productive use
- Enhancing other data with location data
- Structuring and combining location data from multiple sources
- Visualizing data with maps

## Location data

### Taking advantage of location data has been difficult
- Specialized GIS tools and vendors
- Specialized technical knowledge
- Patchy integration with other systems and software

## Location analytics

Increasing recognition of need to bring location data into the mainstream of analytics and data science

- [MIT Sloan Management Review](http://sloanreview.mit.edu/article/location-analytics-bringing-geography-back/)

- [Harvard Business Review](https://hbr.org/2014/03/how-location-analytics-will-transform-retail/)

- [Wired Innovation Insights](http://insights.wired.com/profiles/blogs/location-analytics-where-the-future-will-be)

## Powered by R

> - R provides powerful tools for working with geospatial data

> - R is widely used and widely integrated

> - Many geospatial data manipulation and analysis tasks are within the reach of analysts and data scientists


# Geospatial Data Wrangling with R

## Key R packages for geospatial work

Essential packages

- [**sp**](http://cran.r-project.org/package=sp)
- [**rgdal**](http://cran.r-project.org/package=rgdal)

Other important packages

- [**geosphere**](http://cran.r-project.org/package=geosphere)
- [**rgeos**](http://cran.r-project.org/package=rgeos)
- [**maptools**](http://cran.r-project.org/package=maptools)

## Spatial objects and methods

Classes and methods for spatial data are [defined by the package **sp**](http://cran.r-project.org/web/packages/sp/vignettes/intro_sp.pdf)

- SpatialPoints
- SpatialLines
- SpatialPolygons
- SpatialPointsDataFrame
- SpatialLinesDataFrame
- SpatialPolygonsDataFrame

## Some common tasks

> - Read and write shapefiles
> - Plot spatial data
> - Transform coordinate reference systems
> - Perform spatial overlay (e.g. geofencing)
> - Calculate area, perimeter of polygons
> - Calculate great circle distances
> - Find unions, intersections, differences of polygons

## Load packages

```{r, message=FALSE}
library(sp)
library(rgdal)
library(geosphere)
library(rgeos)
library(maptools)
```

## Read shapefile to SpatialPolygons object

```{r}
shpfile <- "planning_neighborhoods/planning_neighborhoods.shp"
ogrListLayers(shpfile)
neighborhoods <- readOGR(shpfile, "planning_neighborhoods")
```

## Examine SpatialPolygons object

```{r}
class(neighborhoods)
length(neighborhoods)
```

## Plot SpatialPolygons object

```{r}
plot(neighborhoods)
```

## Transform coordinate reference system

```{r}
proj4string(neighborhoods)
bbox(neighborhoods)
```

## Transform coordinate reference system

See [spatialreference.org](http://spatialreference.org/)

```{r}
neighborhoods.xf <- spTransform(
	neighborhoods, 
	CRS=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
)
```
```{r}
proj4string(neighborhoods.xf)
bbox(neighborhoods.xf)
```

## Write shapefile

```{r, eval=FALSE}
shpfile.xf <- "planning_neighborhoods_wgs84.shp"
writeOGR(
  neighborhoods.xf,
  shpfile.xf,
  "planning_neighborhoods_wgs84",
  driver="ESRI Shapefile"
)
```

## Read data with longitude and latitute coordinates

```{r}
restaurants <- read.table(
	file="San Francisco Restaurants.txt",
	sep="\t",
	header=TRUE,
	stringsAsFactors=FALSE,
	comment.char="",
	fill=TRUE,
	quote=""
)
```
Remove rows with missing coordinates
```{r}
location.known <- !is.na(restaurants$longitude) & 
	!is.na(restaurants$latitude)
restaurants <- restaurants[location.known, ]
```

## Examine data

```{r}
nrow(restaurants)
head(restaurants, 3)
```

## Create SpatialPoints object

```{r}
restaurants.sp <- SpatialPoints(
	as.matrix(restaurants[, c("longitude","latitude")]), 
	proj4string=CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
)
```
```{r}
length(restaurants.sp)
bbox(restaurants.sp)
```

## Plot SpatialPoints object

```{r}
plot(restaurants.sp)
```

## Spatial overlay

```{r}
overlay.df <- restaurants.sp %over% neighborhoods.xf
restaurants$neighborhood <- as.character(overlay.df$neighborho)
head(restaurants, 3)
```

## Examine result of spatial overlay

```{r}
sort(table(restaurants$neighborhood), dec=TRUE)
```

## Calculate areas of polygons

```{r}
area <- areaPolygon(neighborhoods.xf) # square meters
area <- area / 2589988 # square miles
sum(area)
names(area) <- neighborhoods.xf@data$neighborho
sort(area, dec=TRUE)
```

## Great circle distance

```{r}
ds <- data.frame(
	City=c("San Francisco", "New York"),
	Longitude=c(-122.416700, -74.005900),
	Latitude=c(37.783300, 40.712700)
)
distHaversine(ds[1, 2:3], ds[2, 2:3], r=3963.17)
```

Argument `r` is the the radius of the earth in the desired units. Use 3963.17 for miles; use 6378137 for meters; use 6378.137 for kilometers.

Also see functions `distCosine`, `distVincentyEllipsoidSphere`, `distVincentyEllipsoid` in package **geosphere**

## Convenience functions

The package **maptools** provides convenience functions for reading and writing shapefiles

```{r}
nc1 <- readShapePoly(
	system.file("shapes/sids.shp", package="maptools")[1],
	proj4string=CRS("+proj=longlat +datum=NAD27")
)
```

Also see `writePolyShape`

## 

```{r}
plot(nc1)
```

## Union, intersection, difference of polygons

Cut North Carolina into four regions by longitude and take unions of counties in the four regions

```{r}
lps <- coordinates(nc1)
ID <- cut(lps[,1], quantile(lps[,1]), include.lowest=TRUE)
reg4 <- gUnaryUnion(nc1, ID)
row.names(reg4)
```

Also see `gUnion`, `gIntersection`, `gDifference` in package **rgeos**

##

```{r}
plot(reg4)
```

## Resources

> - [spatialreference.org](http://spatialreference.org/)
> - [prj2epsg.org](http://prj2epsg.org/)
> - [QGIS](http://www.qgis.org/)
> - [Creating maps in R](https://github.com/Robinlovelace/Creating-maps-in-R)


# Geospatial Analysis with R

## R packages for geospatial analysis

See [CRAN Task View: Analysis of Spatial Data](http://cran.r-project.org/view=Spatial)

## Thank you

[meetup.com/Triangle-useR](http://meetup.com/Triangle-useR/)

[github.com/TriangleR](https://github.com/TriangleR)


